---
title: "Differential expression - APP/MAPT versus WT dirty mice mRNA"
author: "Kimberly Olney, PhD"
date: "January 2026"
output:
  pdf_document: default
params:
  project: APPMAPT (AM)
---

Compare AM versus WT mice, clean versus dirty 

# Set up working enivornment
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = ".")
getwd()
```

```{r libraries, message=FALSE, warning=FALSE}
source("/tgen_labs/jfryer/kolney/dirty_mice/dirty_mouse_AM/scripts/file_paths_and_colours.R")
projectID <- "AM"
```

# Create a counts matrix
This is the raw counts 
```{r raw_counts}
# path to counts files
count_files <-
  file.path(paste0(
    "starAligned/",
    metadata$sample, "_STAR.bam", #  "_", metadata$group,"
    typeOfCount
  ))
# add sample counts name to the counts files
names(count_files) <- paste0(metadata$sample)
# Inspect 
table(metadata$sample)

# Create the counts matrix 
count_data <- data.frame(fread(count_files[1]))[c(1,4)] # forth column for RF reads 
# Loop and read the 4th column
for(i in 2:length(count_files)) {
        count_data <- cbind(count_data, data.frame(fread(count_files[i]))[4])
}
# set gene_id as the row name
row.names(count_data) <- count_data$V1
```

# Mapping metrics
```{r mapping_metrics}
star_metrics <- data.frame(count_data[count_data$V1 %like% "N_", ])
star_metrics$V1 <- as.vector(star_metrics$V1)
melt_star_metrics <- reshape2::melt(star_metrics, id = c("V1"))
ggplot(melt_star_metrics, aes(x = value, fill = V1)) + 
  geom_histogram(position = "identity", bins = 100) + 
  facet_grid(V1 ~ .)  + theme_bw()
```

clean up
```{r}
remove(melt_star_metrics, star_metrics, i, count_files)
```

# Format counts table
```{r counts_table}
# remove star metric information
count_data <- count_data[!grepl("N_", count_data$V1),]
# set gene_id (i.e. V1) as the row name
row.names(count_data) <- count_data$V1
count_data$V1 <- NULL
# set column names to sample counts file name
colnames(count_data) <- metadata$sample

genes <- read.delim("scripts/ensembl_mouse_genes.txt")
```

# Check inputs
```{r check_inputs}
all.equal(rownames(count_data), genes$gene_id) # gene ids match the count rows
all.equal(as.character(colnames(count_data)), as.character(metadata$sample))
```

# Create DGE object
```{r DGE_object}
# create object
dge <- DGEList(counts = count_data,
               samples = metadata,
               genes = genes)

# Inspect
table(dge$samples$group)
table(dge$samples$sample)

# Split by sex
male_samples <- dge$samples$sex == "M"
female_samples <- dge$samples$sex == "F"

dge_male <- dge[, male_samples]
dge_female <- dge[, female_samples]
```

# Set factors and colors 
```{r factor_and_color_set}
dge$samples$group <- factor(dge$samples$group, levels = c("clean_WT", "clean_AM", "dirty_WT", "dirty_AM")) # 
condition_colors <- c("#56B4E9", "#009E73", "#F0E442", "#E69F00")[dge$samples$group]

dge$samples$sex <- factor(dge$samples$sex, levels = c("F", "M"))
sex_colors <- c("purple", "orange")[dge$samples$sex]

dge_male$samples$group <- factor(dge_male$samples$group, levels = c("clean_WT", "clean_AM", "dirty_WT", "dirty_AM")) # 
condition_colors <- c("#56B4E9", "#009E73", "#F0E442", "#E69F00")[dge_male$samples$group]

dge_female$samples$group <- factor(dge_female$samples$group, levels = c("clean_WT", "clean_AM", "dirty_WT", "dirty_AM")) # 
condition_colors <- c("#56B4E9", "#009E73", "#F0E442", "#E69F00")[dge_female$samples$group]
```

# Raw CPM
```{r raw_cpm}
lcpm <- edgeR::cpm(dge$counts, log = TRUE)
lcpm_male <- edgeR::cpm(dge_male$counts, log = TRUE)
lcpm_female<- edgeR::cpm(dge_female$counts, log = TRUE)
```

# Raw MDS
```{r raw_MDS}
# Condition
par(bg = 'white')
plotMDS(
  lcpm,
  top = 100,
  labels = dge$samples$sample,
  cex = 1,
  dim.plot = c(1, 2),
  plot = TRUE,
  col = sex_colors
)
title(expression('Top 100 Genes - Raw (Log'[2] ~ 'CPM)'))
saveToPDF(paste0("results_", projectID, "/both_sexes/MDS/MDS_raw_sex.pdf"), width = 6, height = 6)


par(bg = 'white')
plotMDS(
  lcpm_male,
  top = 100,
  labels = dge_male$samples$sample,
  cex = 1,
  dim.plot = c(1, 2),
  plot = TRUE,
  col = condition_colors
)
title(expression('Top 100 Genes - Raw (Log'[2] ~ 'CPM)'))
saveToPDF(paste0("results_", projectID, "/sex_stratified/MDS/MDS_raw_male.pdf"), width = 6, height = 6)


par(bg = 'white')
plotMDS(
  lcpm_female,
  top = 100,
  labels = dge_female$samples$sample,
  cex = 1,
  dim.plot = c(1, 2),
  plot = TRUE,
  col = condition_colors
)
title(expression('Top 100 Genes - Raw (Log'[2] ~ 'CPM)'))
saveToPDF(paste0("results_", projectID, "/sex_stratified/MDS/MDS_raw_female.pdf"), width = 6, height = 6)
```


# Remove mitochondrial genes
```{r MT_genes}
dim(dge) # inspect
removeMT <- dge$genes$Chr != "chrM"  # true when NOT MT
table(removeMT) # inspect
dge <- dge[removeMT,,keep.lib.sizes = FALSE]
dim(dge)


dim(dge_male) # inspect
removeMT <- dge_male$genes$Chr != "chrM"  # true when NOT MT
table(removeMT) # inspect
dge_male <- dge_male[removeMT,,keep.lib.sizes = FALSE]
dim(dge_male)


dim(dge_female) # inspect
removeMT <- dge_female$genes$Chr != "chrM"  # true when NOT MT
table(removeMT) # inspect
dge_female <- dge_female[removeMT,,keep.lib.sizes = FALSE]
dim(dge_female)
```

# Keep only protein coding genes 
```{r MT_genes}
dim(dge)
table(dge$genes$gene_type)
removeMT <- dge$genes$gene_type == "protein_coding"  # true when NOT MT
dge <- dge[removeMT,,keep.lib.sizes = FALSE]
dim(dge)


dim(dge_male)
table(dge_male$genes$gene_type)
removeMT <- dge_male$genes$gene_type == "protein_coding"  # true when NOT MT
dge_male <- dge_male[removeMT,,keep.lib.sizes = FALSE]
dim(dge_male)


dim(dge_female)
table(dge_female$genes$gene_type)
removeMT <- dge_female$genes$gene_type == "protein_coding"  # true when NOT MT
dge_female <- dge_female[removeMT,,keep.lib.sizes = FALSE]
dim(dge_female)
```

# Library sizes
```{r library}
# before filtering
L <- mean(dge$samples$lib.size) * 1e-6
M <- median(dge$samples$lib.size) * 1e-6
c(L, M)


L_male <- mean(dge_male$samples$lib.size) * 1e-6
M_male <- median(dge_male$samples$lib.size) * 1e-6
c(L_male, M_male)

L_female <- mean(dge_female$samples$lib.size) * 1e-6
M_female <- median(dge_female$samples$lib.size) * 1e-6
c(L_female, M_female)
```

# Filtering 
The filterByExpr() function in the edgeR package determines which genes have a great enough count value to keep.  
We will filter by group_sex.  
```{r filter_counts}
# first filter by expression
dim(dge)
keep.expr <-
  filterByExpr(
    dge,
    group = dge$samples$group, # by disease groups
    min.count = M, # min count of 1 CPM 
  )
dge.filtered <- dge[keep.expr, ,keep.lib.sizes = FALSE]
dim(dge.filtered)
table(dge.filtered$genes$gene_type)



dim(dge_male)
keep.expr <-
  filterByExpr(
    dge_male,
    group = dge_male$samples$group, # by disease groups
    min.count = M_male, # min count of 1 CPM 
  )
dge.filtered.male <- dge_male[keep.expr, ,keep.lib.sizes = FALSE]
dim(dge.filtered.male)
table(dge.filtered.male$genes$gene_type)


dim(dge_female)
keep.expr <-
  filterByExpr(
    dge_female,
    group = dge_female$samples$group, # by disease groups
    min.count = M_female, # min count of 1 CPM 
  )
dge.filtered.female <- dge_female[keep.expr, ,keep.lib.sizes = FALSE]
dim(dge.filtered.female)
table(dge.filtered.female$genes$gene_type)
```

# TMM
For estimating relative RNA production levels from RNA-seq data. 
The TMM method estimates scale factors between samples that can be incorporated 
into currently used statistical methods for DE analysis.
```{r TMM}
# Now, method of trimmed mean of M-values (TMM)
# Calculate scaling factors to convert raw library sizes into effective library sizes.
dge.filtered.norm <- calcNormFactors(dge.filtered, method = "TMM")

# norm factor summary
summary(dge.filtered.norm$samples$norm.factors)
normfactors <- (dge.filtered.norm$samples$norm.factors)

dge.filtered.norm.male <- calcNormFactors(dge.filtered.male, method = "TMM")

# norm factor summary
summary(dge.filtered.norm.male$samples$norm.factors)
normfactors_male <- (dge.filtered.norm.male$samples$norm.factors)

dge.filtered.norm.female <- calcNormFactors(dge.filtered.female, method = "TMM")

# norm factor summary
summary(dge.filtered.norm.female$samples$norm.factors)
normfactors_female <- (dge.filtered.norm.female$samples$norm.factors)
```

# Gene_id to gene_name
Some of the gene_names are duplicated. Our counts matrix has rownames by gene_id which are not duplicated. Since we have now filtered and normalized we can check to see if duplicated gene_names still exist.  If none do, rename rows to gene_name.  If there are duplicates, handle the error and then rename columns.
```{r gene_name}
# check for duplication
table(duplicated(dge.filtered.norm$genes$gene_name))

# replace NA gene names with gene_id first
dge.filtered.norm$genes <- dge.filtered.norm$genes %>%
  mutate(gene_name = coalesce(gene_name, gene_id))

# replace duplicated gene names after the first occurrence with gene_id
dge.filtered.norm$genes <- dge.filtered.norm$genes %>%
  mutate(
    gene_name = if_else(
      duplicated(gene_name),
      gene_id,
      gene_name
    )
  )

# check duplication
table(duplicated(dge.filtered.norm$genes$gene_name))


# replace NA gene names with gene_id first
dge.filtered.norm.male$genes <- dge.filtered.norm.male$genes %>%
  mutate(gene_name = coalesce(gene_name, gene_id))

# replace duplicated gene names after the first occurrence with gene_id
dge.filtered.norm.male$genes <- dge.filtered.norm.male$genes %>%
  mutate(
    gene_name = if_else(
      duplicated(gene_name),
      gene_id,
      gene_name
    )
  )

# replace NA gene names with gene_id first
dge.filtered.norm.female$genes <- dge.filtered.norm.female$genes %>%
  mutate(gene_name = coalesce(gene_name, gene_id))

# replace duplicated gene names after the first occurrence with gene_id
dge.filtered.norm.female$genes <- dge.filtered.norm.female$genes %>%
  mutate(
    gene_name = if_else(
      duplicated(gene_name),
      gene_id,
      gene_name
    )
  )
```

# Clean up
```{r cleanup}
rm(count_data, lcpm, keep.expr, normfactors, removeMT, L, M, typeOfCount)
```


# Voom 
voom transform counts to use for variancePartition  
```{r voom, warning=FALSE}
formula <- (~ 0 + group)
counts <- as.data.frame(dge.filtered.norm$counts)
info <- as.data.frame(dge.filtered.norm$samples)
voom_with_weights <-
  variancePartition::voomWithDreamWeights(
    counts = counts,
    formula = formula,
    data = info,
    BPPARAM = BiocParallel::SnowParam(8),
    plot = TRUE
  )
path <- paste0(paste0("results_", projectID, "/both_sexes/voom/voom_raw"))
saveToPDF(paste0(path, ".pdf"), width = 6, height = 4)
voomCounts <- voom_with_weights$E
rm(voom_with_weights)


formula <- (~ 0 + group)
counts_male <- as.data.frame(dge.filtered.norm.male$counts)
info_male <- as.data.frame(dge.filtered.norm.male$samples)
voom_with_weights_male <-
  variancePartition::voomWithDreamWeights(
    counts = counts_male,
    formula = formula,
    data = info_male,
    BPPARAM = BiocParallel::SnowParam(8),
    plot = TRUE
  )
path <- paste0(paste0("results_", projectID, "/sex_stratified/voom/voom_raw_male"))
saveToPDF(paste0(path, ".pdf"), width = 6, height = 4)
voomCounts <- voom_with_weights_male$E
rm(voom_with_weights_male)


formula <- (~ 0 + group)
counts_female <- as.data.frame(dge.filtered.norm.female$counts)
info_female <- as.data.frame(dge.filtered.norm.female$samples)
voom_with_weights_female <-
  variancePartition::voomWithDreamWeights(
    counts = counts_female,
    formula = formula,
    data = info_female,
    BPPARAM = BiocParallel::SnowParam(8),
    plot = TRUE
  )
path <- paste0(paste0("results_", projectID, "/sex_stratified/voom/voom_raw_female"))
saveToPDF(paste0(path, ".pdf"), width = 6, height = 4)
voomCounts <- voom_with_weights_female$E
rm(voom_with_weights_female)
```


# Design matrix
```{r design_matrix}
design <- model.matrix( ~ 0 + dge.filtered.norm$samples$group + dge.filtered.norm$samples$sex + dge.filtered.norm$samples$RIN)
colnames(design) <- c("clean_WT", "clean_AM", "dirty_WT", "dirty_AM", "sex", "RIN")
# Inspect
design

design_male <- model.matrix( ~ 0 + dge.filtered.norm.male$samples$group + dge.filtered.norm.male$samples$RIN)
colnames(design_male) <- c("clean_WT", "clean_AM", "dirty_WT", "dirty_AM", "RIN")
# Inspect
design_male

design_female <- model.matrix( ~ 0 + dge.filtered.norm.female$samples$group + dge.filtered.norm.female$samples$RIN)
colnames(design_female) <- c("clean_WT", "clean_AM", "dirty_WT", "dirty_AM", "RIN")
# Inspect
design_female
```

# Voom
When the library sizes are quite variable between samples, then the voom approach is theoretically more powerful than limma-trend. 
The voom method estimates the mean-variance relationship of the log-counts.
Generates a precision weight for each observation and enters these into the limma empirical Bayes analysis pipeline.
```{r voom_covariates}
v <- voomWithQualityWeights(
  dge.filtered.norm,  # our DGEList object
  design, # our design matrix
  plot = TRUE # should mean-variance trend & sample-specific weights be displayed?
  )

v_male <- voomWithQualityWeights(
  dge.filtered.norm.male,  # our DGEList object
  design_male, # our design matrix
  plot = TRUE # should mean-variance trend & sample-specific weights be displayed?
  )

v_female <- voomWithQualityWeights(
  dge.filtered.norm.female,  # our DGEList object
  design_female, # our design matrix
  plot = TRUE # should mean-variance trend & sample-specific weights be displayed?
  )
```

# Contrast plot
```{r contrasts}
# fits linear model for each gene given a series of arrays
fit_male <- lmFit(v_male, design_male) # voom_cov
coef.fit_male <- fit_male$coefficients

contrasts_male <- makeContrasts(
  clean_AM_vs_clean_WT = clean_AM - clean_WT,
  dirty_AM_vs_dirty_WT = dirty_AM - dirty_WT,
  dirty_WT_vs_clean_WT = dirty_WT - clean_WT,
  dirty_AM_vs_clean_AM = dirty_AM - clean_AM,
  levels = colnames(design_male))
head(contrasts_male)

# save contrast names
allComparisons_male <- colnames(contrasts_male)
allComparisons_male # check

# run contrast analysis
vfit_male <- contrasts.fit(fit_male, contrasts = contrasts_male)

# Compute differential expression based on the empirical Bayes moderation of the standard errors towards a common value.
# The logCPM values can then be used in any standard limma pipeline, using the trend=TRUE
veBayesFit_male <- eBayes(vfit_male, trend = TRUE, robust=TRUE)
plotSA(veBayesFit_male, main = "Final Model: Mean-variance Trend")

fit_female <- lmFit(v_female, design_female) # voom_cov
coef.fit_female <- fit_female$coefficients

contrasts_female <- makeContrasts(
  clean_AM_vs_clean_WT = clean_AM - clean_WT,
  dirty_AM_vs_dirty_WT = dirty_AM - dirty_WT,
  dirty_WT_vs_clean_WT = dirty_WT - clean_WT,
  dirty_AM_vs_clean_AM = dirty_AM - clean_AM,
  levels = colnames(design_female))
head(contrasts_female)

# save contrast names
allComparisons_female <- colnames(contrasts_female)
allComparisons_female # check

# run contrast analysis
vfit_female <- contrasts.fit(fit_female, contrasts = contrasts_female)

# Compute differential expression based on the empirical Bayes moderation of the standard errors towards a common value.
# The logCPM values can then be used in any standard limma pipeline, using the trend=TRUE
veBayesFit_female <- eBayes(vfit_female, trend = TRUE, robust=TRUE)
plotSA(veBayesFit_female, main = "Final Model: Mean-variance Trend")
```

# DEGs summary
```{r DGE_summary}
qval <- 0.05
lfc.cutoff <- 0

sumTable_male <- 
  summary(decideTests(
    veBayesFit_male,  # object
    adjust.method = "BH", # by default the method = "separate"
    p.value = qval,
    lfc = lfc.cutoff  # numeric, minimum absolute log2-fold change required
  ))

print(paste0("male q < ", qval, " & absolute log2-fold change > ", lfc.cutoff))
sumTable_male

sumTable_female <- 
  summary(decideTests(
    veBayesFit_female,  # object
    adjust.method = "BH", # by default the method = "separate"
    p.value = qval,
    lfc = lfc.cutoff  # numeric, minimum absolute log2-fold change required
  ))

print(paste0("female q < ", qval, " & absolute log2-fold change > ", lfc.cutoff))
sumTable_female

genes_relevant <- dge.filtered.norm$genes
```

# DEG heatmap 
```{r DEG_heatmap}
sumTable_male <- as.data.frame(sumTable_male)
colnames(sumTable_male) <- c("direction", "comparison", "count")
sumTable_male <- sumTable_male[!sumTable_male$direction == "NotSig",]
sumTable_male <- sumTable_male |> 
  pivot_wider(names_from = comparison, values_from = count)
row_names_to_keep <- sumTable_male$direction
sumTable_male <- sumTable_male[, -which(names(sumTable_male) == "direction")]
sumTable_male <- t(sumTable_male)
colnames(sumTable_male) <- row_names_to_keep
df <- as.data.frame(sumTable_male)

pdf(paste0("results_", projectID, "/sex_stratified/DEGs/DEGs_heatmap_male.pdf"), height = 6, width = 6)
pheatmap::pheatmap(df,
                   main = paste0("q < ", qval, " & |log2FC| > ", lfc.cutoff),
                   treeheight_row = 0,
                   treeheight_col = 0,
                   color = rev(natparks.pals("WindCave")),
                   cluster_rows = FALSE,
                   display_numbers = round(df, digits = 0),
                   fontsize_number = 12,
                   number_color = ifelse(df > 400, "white", "black"),
                   angle_col = "0",
                   breaks = c(seq(0, 500, length.out = length(rev(natparks.pals("WindCave"))))))
dev.off()


sumTable_female <- as.data.frame(sumTable_female)
colnames(sumTable_female) <- c("direction", "comparison", "count")
sumTable_female <- sumTable_female[!sumTable_female$direction == "NotSig",]
sumTable_female <- sumTable_female |> 
  pivot_wider(names_from = comparison, values_from = count)
row_names_to_keep <- sumTable_female$direction
sumTable_female <- sumTable_female[, -which(names(sumTable_female) == "direction")]
sumTable_female <- t(sumTable_female)
colnames(sumTable_female) <- row_names_to_keep
df <- as.data.frame(sumTable_female)

pdf(paste0("results_", projectID, "/sex_stratified/DEGs/DEGs_heatmap_female.pdf"), height = 6, width = 6)
pheatmap::pheatmap(df,
                   main = paste0("q < ", qval, " & |log2FC| > ", lfc.cutoff),
                   treeheight_row = 0,
                   treeheight_col = 0,
                   color = rev(natparks.pals("WindCave")),
                   cluster_rows = FALSE,
                   display_numbers = round(df, digits = 0),
                   fontsize_number = 12,
                   number_color = ifelse(df > 400, "white", "black"),
                   angle_col = "0",
                   breaks = c(seq(0, 500, length.out = length(rev(natparks.pals("WindCave"))))))
dev.off()
```


# Output DEG tables
```{r output_DEG_tables}
coef <- 1

for (i in allComparisons) {
  vTopTableAll <- topTable(
    veBayesFit, 
    coef = coef,  
    n = Inf, 
    p.value = 1,
    lfc = 0, 
    sort.by = "P", 
    genelist = genes_relevant, 
    confint = TRUE # column of confidence interval 
    )
    saveRDS(vTopTableAll, file = 
            paste0("rObjects/gene_tables/", i,"_gene_table.rds"))
  path <- paste0("results_", projectID, "/both_sexes/DEGs/", i, "_q1.00.txt", sep = "") 
  write.table(vTopTableAll, path, sep = "\t", row.names = FALSE, quote = FALSE)
  
  # p < 0.05, log2fc > 1
  vTopTableSig <-
    topTable( 
      veBayesFit,  
      coef = coef,  
      n = Inf, 
      p.value = qval,
      lfc = lfc.cutoff,
      genelist = genes_relevant, 
      confint = TRUE)
  path <- paste0("results_", projectID, "/both_sexes/DEGs/", i, "_q", qval, "_lfc", lfc.cutoff,".txt", sep = "") 
  write.table(vTopTableSig, path, sep = "\t", row.names = FALSE, quote = FALSE)
  # increment 
  coef <- coef + 1
}
remove(coef, vTopTableAll, vTopTableSig)
```
